Refunctionalization of Abstracting Abstract Machines (Functional Pearl)

### 2018-03-16 paper.tex
Fixed spacing, typos

boardly speaking, AAM and ADI they are both abstract interpreters.

@@

\item A new understanding of pushdown control-flow analysis by 
  refunctionalization.
  We present an abstract interpreter obtained from refunctionalizing
  an small-step abstracting abstract machine. 
  This abstract interpreter naturally enjoys pushdown control-flow property
  because the refunctionalization forces us to sequentialize the 
  abstract exectuion.
  We also analyze its correctness, soundness and termination.

@@

In section \ref{background}, we first describe the target language we will
analyze and a concrete CESK machine as small-step operational semantics for it.
Then we review abstracting abstract machines (AAM) as an abstract interpretation
counterpart of CESK machines. In section \ref{linear}, we convert the
non-deterministic choices of closures to another layer of continuations,
therefore linearize the abstract execution.
Section \ref{fusing} fuses the driver function and step function to one single
function.
In section \ref{disen}, we further identify the first-order data type representations of
continuations, and disentangle the code blocks for dispatching data types to separate
functions.
Section \ref{refunc} represents the refunctionalized abstract abstract machines,
In this part, the continuations represented by data type trun to higher-order
functions.
In Section \ref{directstyle}, we transform the refunctionalized AAM to two
different direct style, one is implemented with monads, the other is using
delimited controls.

@@

\subsubsection{Allocating Addresses}
Up to now, we did not describe about allocating adresses in stores and the time stamp @Time@.
But in abstract interpretation, they are key ingredients to achieve 
analyses with different sensitivities, as well as to a finite state space analysis\cite{Gilray:2016:ACP:2951913.2951936}.
To effectively approximate the runtime behavior, we introduce a finite 
program contour @time@ that encodes the program execution history.
Here, we use a list of execution contexts (expressions) to represent, and
as we will see in section \ref{kcfainst}, by applying different @tick@
functions on time stamp we may obtain a family of analyses.

@@

Since dereferencing an address becomes non-deterministic, our @atomicEval@
function also truns to be non-deterministic. Given an atomic expression @e@,
@atomicEval@ returns a set of storable values, i.e., closures, to the caller.
If the expression is just a @lambda@ term, then the returned set is a singleton.

@@

A histroy of the last $k$ call sites is used as a finite program contour in
$k$-call-sensitive analysis.
The history is represented as a list of expressions and embedded
in the allocated addresses.

@@

If we instantiate $k$ to be $0$, the history degenerates to an empty list,
and we obtain a monovariant analysis, which does
not differentiate values at different call sites. In this case, the address
space collapses to the space of variable names.
Note that regarding the ambiguitiy in $k$-CFA\cite{Gilray:2016:ACP:2951913.2951936},
the code here actually implements call+return sensitivity.

@@

In Danvy's paper \textit{Defunctionalized Interpreters for Programming
  Languages}, he metioned for deterministic languages, a reduction semantics
is a structural operational semantics in continuation style, where the reduction
context is a defunctionalized continuation \cite{Danvy:2008:DIP:1411204.1411206}.
Here we are facing a problem:
the underlying semantics of AAM is fundamentally non-deterministic.
But the evaluation context (the @Frame@ in our program) is not
non-deterministic, and the work list @todo@ actually implicitly handles the non-determinism,
if we simply refunctionalize the frames to functions, it would not helps toward
abstract definitional interpreters.

In order to address this problem, our first step of tramsformations
is to linearize all the non-deterministic choices.
This step turns the @step@ function to be deterministic.
Likewise the @Frame@ saves the information of the caller in concrete executions,
we introduce another layer of controls, and define a case class @NDCont@
that saves the information at a fork point when we have multiple target closures.
We also add a new field @ndk@ to the definition of state which is called
@NDState@ now.
For the rest of the paper, to differentiate the two continuations, we would call
the first one as normal continuation, and the second one as non-deterministic
continuations.

@@

Fusing is a tramsformation that combines function @step@ and @drive@ to one.
The fused function @drive_step@ basically formed by merging function
@step@ into function @drive@.
@drive_step@ takes a @NDState@ and a set of explored states as argument,
and returns a set of reachable states once terminates.

@@

Now we have a single function to perform both abstract evaluation and intermediate
states collecting. Given @inject(e)@ as the inital state and an empty set as
initial reached set, we can easily define the entrance function @analyze@ as follow.

@@

The above code shows the skeleton of function @drive_step@.
In the first case of pattern matching, the code for constructing new
environment, store and frames is omitted.
At the end of this case, we invoke @drive_step@ recursively.
The code for the second case is replaced by a function call to @continue@.
Now a vague shape of interpreer in continuation style is emerged.

@@

In this section, several other notable changes have been made:
\begin{itemize}
\item We do not have the concept of @state@ (or @NDState@) anymore.
  The components of the states are lifted as arguments of the evaluation
  function.
\item Since our final target is abstract definitional interpreters while still
non-deterministic, start from this section, our abstract semantic artifact
returns a set of final values insteaf of collected states.
\item As we are getting close to the
abstract definitional interperters, the name of @drive_step@ is updated to
@aeval@ (for \emph{abstract eval}).
\end{itemize}

For clearity, in this section, we first show how to refunctionalize the normal continuation and
non-deterministic continuation without collecting semantics.
Then adding an appropriate caching algorithm to make sure the analysis will
terminate at some fixed-point is just changing the program to cache-passing style.
Finally, we examine how computable pushdown control-flow analysis is established
through refunctionalization and caching.

@@

\textbf{First Version of Evaluation Function.}
Since our first version of @aeval@ just implements the non-deterministic
evaluation, and given the defined @nd@, the evaluation function @aeval@ can be easily
defined. For the case that @e@ is a @let@ expression,
the overall idea is to perform a two step depth-first evluation. \todo{refine}

\begin{lstlisting}
case Let(x, App(f, ae), e) =>
  val closures = atomicEval(f, env, store).asInstanceOf[Set[Clos]]
  nd[Clos, Ans](closures, Set[VS](), { case (clos, accouter, closnd) =>
    val Clos(Lam(v, body), c_env) = clos
    val baddr = allocBind(v, new_time)
    val new_env = c_env + (v -> baddr)
    val new_store = store.update(baddr, atomicEval(ae, env, store)) 
    aeval(body, new_env, new_store, new_time, (bodyvss: Set[VS]) => {
      nd[VS, Ans](bodyvss, Set[VS](), { case (vs, accinner, bdnd) => 
        val VS(vals, time, store) = vs
        val new_env = env + (x -> baddr)
        val new_store = store.update(baddr, vals)
        aeval(e, new_env, new_store, time, 
              /* accumulate the values of one path, and call next body value*/
              (evss: Ans) => bdnd(accinner ++ evss)) 
      },
      /* accumulate the values of pathes to accouter, and call next closure */
      (evss: Ans) => closnd(evss ++ accouter)) 
    })
  }, 
  continue)
\end{lstlisting}

The outer function call @nd@ evaluates over the set of possible closures of
@f@. For each closures, we go into its body with new environment and new store.
Remember that the function @aeval@ returns a set of @VS@ to its continuation,
so the continuation argument @bodyvss@ is a set of @VS@ that represents values
from multiple computation pathes when evaluating this single @body@ expression.

@@

This paper is inspired a lot from \citeauthor{darais2017abstracting}'s work.
Instead of just showing the final form of abstract definitional interpreters,
We reveal that refunctionalization plays an important role for inheriting
stack from the meta-language.
The target of the series of transformations in this paper is abstract definitional
interpreters, and we additionaly demonstrate abstract interpreters written in both
monadic- and direct-style.
The difference is ours work for A-Normal Form $\lambda$-calculus instead of plain
$\lambda$-calculus, our work also can be easily adopted to plain
$\lambda$-calculus.

>>>>

broadly speaking, AAM and ADI they are both abstract interpreters.

@@

\item A new understanding of pushdown control-flow analysis by
  refunctionalization.
  We present an abstract interpreter obtained from refunctionalizing
  an small-step abstracting abstract machine.
  This abstract interpreter naturally enjoys pushdown control-flow property
  because the refunctionalization forces us to sequentialize the
  abstract execution.
  We also analyze its correctness, soundness and termination.

@@

In section \ref{background}, we first describe the target language we will
analyze and a concrete CESK machine as small-step operational semantics for it.
Then we review abstracting abstract machines (AAM) as an abstract interpretation
counterpart of CESK machines. In section \ref{linear}, we convert the
non-deterministic choices of closures to another layer of continuations,
therefore linearize the abstract execution.
Section \ref{fusing} fuses the driver function and step function to one single
function.
In section \ref{disen}, we further identify the first-order data type representations of
continuations, and disentangle the code blocks for dispatching data types to separate
functions.
Section \ref{refunc} represents the refunctionalized abstract abstract machines,
In this part, the continuations represented by data type turn to higher-order
functions.
In Section \ref{directstyle}, we transform the refunctionalized AAM to two
different direct style, one is implemented with monads, the other is using
delimited controls.

@@

\subsubsection{Allocating Addresses}
Up to now, we did not describe about allocating addresses in stores and the time stamp @Time@.
But in abstract interpretation, they are key ingredients to achieve
analyses with different sensitivities, as well as to a finite state space analysis\cite{Gilray:2016:ACP:2951913.2951936}.
To effectively approximate the runtime behavior, we introduce a finite
program contour @time@ that encodes the program execution history.
Here, we use a list of execution contexts (expressions) to represent, and
as we will see in section \ref{kcfainst}, by applying different @tick@
functions on time stamp we may obtain a family of analyses.

@@

Since dereferencing an address becomes non-deterministic, our @atomicEval@
function also turns to be nondeterministic. Given an atomic expression @e@,
@atomicEval@ returns a set of storable values, i.e., closures, to the caller.
If the expression is just a @lambda@ term, then the returned set is a singleton.

@@

A history of the last $k$ call sites is used as a finite program contour in
$k$-call-sensitive analysis.
The history is represented as a list of expressions and embedded
in the allocated addresses.

@@

If we instantiate $k$ to be $0$, the history degenerates to an empty list,
and we obtain a monovariant analysis, which does
not differentiate values at different call sites. In this case, the address
space collapses to the space of variable names.
Note that regarding the ambiguity in $k$-CFA\cite{Gilray:2016:ACP:2951913.2951936},
the code here actually implements call+return sensitivity.

@@

In Danvy's paper \textit{Defunctionalized Interpreters for Programming
  Languages}, he mentioned for deterministic languages, a reduction semantics
is a structural operational semantics in continuation style, where the reduction
context is a defunctionalized continuation \cite{Danvy:2008:DIP:1411204.1411206}.
Here we are facing a problem:
the underlying semantics of AAM is fundamentally non-deterministic.
But the evaluation context (the @Frame@ in our program) is not
non-deterministic, and the work list @todo@ actually implicitly handles the non-determinism,
if we simply refunctionalize the frames to functions, it would not helps toward
abstract definitional interpreters.

In order to address this problem, our first step of transformations
is to linearize all the non-deterministic choices.
This step turns the @step@ function to be deterministic.
Likewise the @Frame@ saves the information of the caller in concrete executions,
we introduce another layer of controls, and define a case class @NDCont@
that saves the information at a fork point when we have multiple target closures.
We also add a new field @ndk@ to the definition of state which is called
@NDState@ now.
For the rest of the paper, to differentiate the two continuations, we would call
the first one as normal continuation, and the second one as non-deterministic
continuations.

@@

Fusing is a transformation that combines function @step@ and @drive@ to one.
The fused function @drive_step@ basically formed by merging function
@step@ into function @drive@.
@drive_step@ takes a @NDState@ and a set of explored states as argument,
and returns a set of reachable states once terminates.

@@

Now we have a single function to perform both abstract evaluation and intermediate
states collecting. Given @inject(e)@ as the initial state and an empty set as
initial reached set, we can easily define the entrance function @analyze@ as follow.

@@

The above code shows the skeleton of function @drive_step@.
In the first case of pattern matching, the code for constructing new
environment, store and frames is omitted.
At the end of this case, we invoke @drive_step@ recursively.
The code for the second case is replaced by a function call to @continue@.
Now a vague shape of interpreter in continuation style is emerged.

@@

In this section, several other notable changes have been made:
\begin{itemize}
\item We do not have the concept of @state@ (or @NDState@) anymore.
  The components of the states are lifted as arguments of the evaluation
  function.
\item Since our final target is abstract definitional interpreters while still
non-deterministic, start from this section, our abstract semantic artifact
returns a set of final values instead of collected states.
\item As we are getting close to the
abstract definitional interpreters, the name of @drive_step@ is updated to
@aeval@ (for \emph{abstract eval}).
\end{itemize}

For clarity, in this section, we first show how to refunctionalize the normal continuation and
non-deterministic continuation without collecting semantics.
Then adding an appropriate caching algorithm to make sure the analysis will
terminate at some fixed-point is just changing the program to cache-passing style.
Finally, we examine how computable pushdown control-flow analysis is established
through refunctionalization and caching.

@@

\textbf{First Version of Evaluation Function.}
Since our first version of @aeval@ just implements the non-deterministic
evaluation, and given the defined @nd@, the evaluation function @aeval@ can be easily
defined. For the case that @e@ is a @let@ expression,
the overall idea is to perform a two step depth-first evaluation. \todo{refine}

\begin{lstlisting}
case Let(x, App(f, ae), e) =>
  val closures = atomicEval(f, env, store).asInstanceOf[Set[Clos]]
  nd[Clos, Ans](closures, Set[VS](), { case (clos, accouter, closnd) =>
    val Clos(Lam(v, body), c_env) = clos
    val baddr = allocBind(v, new_time)
    val new_env = c_env + (v -> baddr)
    val new_store = store.update(baddr, atomicEval(ae, env, store))
    aeval(body, new_env, new_store, new_time, (bodyvss: Set[VS]) => {
      nd[VS, Ans](bodyvss, Set[VS](), { case (vs, accinner, bdnd) =>
        val VS(vals, time, store) = vs
        val new_env = env + (x -> baddr)
        val new_store = store.update(baddr, vals)
        aeval(e, new_env, new_store, time,
              /* accumulate the values of one path, and call next body value*/
              (evss: Ans) => bdnd(accinner ++ evss))
      },
      /* accumulate the values of paths to accouter, and call next closure */
      (evss: Ans) => closnd(evss ++ accouter))
    })
  },
  continue)
\end{lstlisting}

The outer function call @nd@ evaluates over the set of possible closures of
@f@. For each closures, we go into its body with new environment and new store.
Remember that the function @aeval@ returns a set of @VS@ to its continuation,
so the continuation argument @bodyvss@ is a set of @VS@ that represents values
from multiple computation pates when evaluating this single @body@ expression.

@@

This paper is inspired a lot from \citeauthor{darais2017abstracting}'s work.
Instead of just showing the final form of abstract definitional interpreters,
We reveal that refunctionalization plays an important role for inheriting
stack from the meta-language.
The target of the series of transformations in this paper is abstract definitional
interpreters, and we additionally demonstrate abstract interpreters written in both
monadic- and direct-style.
The difference is ours work for A-Normal Form $\lambda$-calculus instead of plain
$\lambda$-calculus, our work also can be easily adopted to plain
$\lambda$-calculus.
